Effects of bariatric surgery on resting-state functional connectivity of the reward and default mode network: a pre-registered analysis

### 2021-03-09 manuscript/draft_rsfmri.Rmd
minor changes

---
title: "Effects of bariatric surgery loss on resting-state functional connectivity of the reward and default mode network: a pre-registered analysis"
author:
- Hannah Sophie Heinrichs* [1]
- Frauke Beyer* [1][2]
- Evelyn Medawar [1]
- Kristin Prehn [3][4]
- Jürgen Ordemann [7][8]
- Agnes Floel [5][6]
- A. Veronica Witte [1][2][9]
date: 1 Max-Planck-Institute for Human Cognitive and Brain Sciences, Leipzig \newline
  2 CRC 1052 "Obesity Mechanisms", Subproject A1, University of Leipzig \newline 3 Department of Neurology and NeuroCure Clinical Research Center, Charité University Medicine, Berlin \newline 4 Department of Psychology, Medical School Hamburg, Hamburg \newline 5 Department of Neurology, University of Greifswald, Greifswald \newline 6 German Center for Neurodegenerative Diseases, Standort Rostock/Greifswald, Greifswald \newline 7 Center for Bariatric and Metabolic Surgery, Charité University Medicine, Berlin \newline 8 Center for Bariatric and Metabolic Surgery, Vivantes Clinic Spandau, Berlin \newline 9 Day Clinic for Cognitive Neurology, University Clinic Leipzig
output:
  bookdown::pdf_book:
    extra_dependencies: "subfig"
    fig_caption: yes
    number_sections: yes
    toc: yes
bibliography: "bibliography.bib"
editor_options:
  chunk_output_type: console
header-includes:
- \usepackage{setspace}
- \onehalfspacing
indent: true
---

@@

Obesity imposes serious health risks and is consistently associated with differences in brain activity, including resting-state functional connectivity. Effective treatment can be achieved by bariatric surgery, yet its effects on functional connectivity of brain networks involved in reward processingand intereoception are still under debate. In this pre-registered analysis, we aimed to determine the effects of bari-atric surgery and weight loss on functional connectivity in the RN and default mode network (DMN. 
Therefore, 33 bariatric surgery patients (males/females, BMI, age etc) and 15 obese waiting-list control patients (males/females, BMI, age etc) underwent MRI at baseline (before surgery/waiting period), and after 6 and 12 months. We pre-registered (https://osf.io/f8tpn and https://osf.io/59bh7/) the whole-brain time-by-group interaction on network connectivity, assessed between baseline and followup and for two denoising pipelines. We also pre-registered a time-by-group interaction analysis of aggregated FC within networks.  In exploratory analyses, we investigated all three time points and assessed the effects of weight loss and head motion. 
According to whole-brain marginal models, bariatric surgery compared to waiting did not significantly affect functional connectivity of the the reward and DMN networks (FWE p > 0.05). There also was no interaction effect on within network FC in pre-registered analysis of aggregated FC. Exploratory analyses also did not show an interaction effect for three time points, but when adding differences in head motion within and between individuals using mean framewise displacement, surgery-related BMI decrease and higher average head motion resulted in stronger functional connectivity in medial posterior frontal regions within the RN. 
This pre-registered analysis of longitudinal neuroimaging data did not support a strong effect of bariatric surgery, compared to waiting, on reward and DMN network functional connectivity after 6 months in obese patients. Exploratory analyses indicated that differences in head motion might have confounded effects of weight loss on reward-related functional connectivity, yet definite conclusions remain difficult. Future studies with a more rigorous control of within-scanner head motion during data acquisition are warranted to further understand the effects of obesity surgery on functional connectivity measures.

@@

Obesity is a worldwide health issue, entailing huge personal and societal costs. Excess amount of body fat not only affects cardiovascular and metabolic health, but also increases the risk for cognitive decline and dementia later in life (hier eine Ref). Conservative treatment options including behavioral therapy often do not yield the desired weight loss, especially in patients with very high BMI ($> 35$ kg/m$^2$). Here, bariatric surgery is a viable option to rapidly induce weight loss and improve glycemic status. RYBG is the most invasive, yet most effective surgical pro-cedure, but other techniques like vertical sleeve gastrectomy (VSG) and gastric banding (GB) are also common [@Berthoud_2011]. 
Bariatric surgery induces weight loss  in part by malabsorption of nutrients, especially fat, and reduced digestion efficiency. More importantly, weight loss following the surgical procedures, and especially RYGB, is related to profound alterations in appetite, food perception and the central regulation of food intake [@Mulla_2017; @Brutman_2019]. 
Precise mechanisms how bariatric surgery leads to altered appetitive and reward signaling are yet to be elucidated. Resting-state functional magnetic resonance imaging (rsfMRI) is a technique to investigate the dynamic organization of the brain. Functional connectivity (FC) networks, i.e. brain regions with correlated neural activity over time, are in anatomical correspondence with specific brain networks involved in cognitive processes, including attention and executive control [@Smith_2009]. The reward network (RN), processing hedonic value and internal motivation, and the default mode network (DMN), a higher-order network, involved in interoception and governing shifts between external-internal processes,  are promising candidates to mediate altered central regulation of food intake after bariatric surgery.
The RN (RN), comprises the ventromedial prefrontal cortex (vmPFC), the striatum, the amygdala and the anterior insula  [@Odoherty_2004;@Liu_2011]. These brain regions have been found to guide food valuation processes and decision-making in humans [@Bartra_2013; @Hare_2011; @Hutcherson_2012; @Schmidt_2018]. Frequently, obesity has been associated with hyperactivation of RN regions during anticipation of (high-caloric) food cues, and in contrast, reduced activation to actual taste of these foods [@Devoto_2018; @Garcia_2014; @Meng_2020; @Stoeckel_2009; though see @Morys_2020]. resting-state rsfMRI studies also shown increased local FC of re-ward network regions, i.e. NAcc, vmPFC, putamen, insula [@Contreras_2017; @Coveleskie_2015;@Hogenkamp_2016], and altered connectivity with salience, homeostatic and sensorimotor networks [@Lips_2014;  @Wijngaarden_2015]. In bariatric surgery patients, connectivity  within the RN (e.g. putamen, OFC) might be normalized by the surgery, however the evidence is limited due to a lack of longitudinal obese control groups  [@Duan_2020; @Wiemerslage_2016]. Possibly, a reconfiguration of the fronto-striatal brain networks could emerge from altered gut signaling , e.g. changes in ghrelin levels, via hypothalamic-striatal projections [@Li_2019; @Karra_2013; but @Zoon_2018] . 
The DMN includes the posterior cingulate cortex (PCC)/precuneus, the medial prefrontal cortex (mPFC) and the inferior and lateral parietal cortex (approx. BA 39) [@Raichle_2015]. Higher BMI and obesity have been associated with a pattern of decreased within DMN FC and increased FC of DMN regions to other networks, i.e. salience and sensory networks in several resting-state and task-based rsfMRI studies [@Beyer_2017; @Kullmann_2011; @Chao_2018; @Wijngaarden_2015; @Borowitz_2020; @Ding_2020; @Doucet_2017; @Sadler_2018]. Studies in bariatric surgery patients have pointed to a normalization of the connectivity between DMN and cognitive control and salience brain regions after the procedure, yet no study has included a longitudinal control group [@Li_2018; @Olivo_2017; @Frank_2013].

@@

Most rsfMRI studies in obesity have reported an association of higher BMI and higher head motion during MRI [@Beyer_2017;@Hodgson_2016]. Head motion is a major confounder in rsfMRI as it increases short-range and disrupts long-range connectivity FC. In this sample, we previously reported a group-by-time interaction on head motion [@Beyer_2020]. Here, we thus aimed to rigorously control for motion-related variance. 
In this pre-registered study, our confirmatory hypotheses were:

@@

Participants were recruited from the Center for Bariatric and Metabolic Surgery at the Charité University Medicine Berlin. Inclusion criteria were, in accordance with German guidelines for bariatric surgery, a failure of conservative obesity treatment and either (1) a BMI $> 40$ km/m$^2$ or (2) a BMI $> 35$ kg/m$^2$ and at least one typical co-morbidity (e.g. type-2 diabetes, hypertension, non-alcoholic fatty liver disease)[@Mechanick_2013]. Participants were aged between $18$ and $70$ years and had no history of cancer, chronic inflammatory disease and addiction, other severe untreated diseases, brain pathologies identified in the MRI scan or cognitive impairments (MMSE score < $34$).
 In total, $51$ participants out of the originally enrolled $69$ subjects received MRI, and we analyzed a total of $101$ data points. Five data points of three subjects had to be excluded due to bad anatomical image quality (see below).  
<!-- DESCRIPTIVE STATISTICS -->
The final sample entailed `r nrow(subj_df)` morbidly obese individuals (`r nrow(filter(subj_df,Sex == "f"))` females; aged `r round(mean(subj_df$Age_BL),2)` $\pm$ `r round(sd(subj_df$Age_BL),2)` SD years (range `r min(subj_df$Age_BL)`-`r max(subj_df$Age_BL)`). Participants either underwent surgery (n = `r cs["Sum","IG"]`, `r cs["f","IG"]` female) or were waiting list controls (n = `r cs["Sum","KG"]`, `r cs["f","KG"]` female), who waited for their health insurance's approval to undergo surgery. Measures were taken at baseline (BL), $6$ (FU1) and $12$ (FU2) months post-surgery/baseline appointment to capture both phases of rapid weight-loss and maintenance [@Maciejewski_2016]. 
Analyses were performed on all participants that provided at least one data point of rsfMRI data. `r sum(freqtable$Freq == 3)` participants had complete data, `r sum(freqtable$Freq == 2)` provided data for two time points and `r sum(freqtable$Freq == 1)` for one time point (for more details see Table S1). The pre-planned analysis of changes from baseline to follow-up included $24$ participants with both time points, in total $72$ data points.

@@

Participants arrived in the morning (between 07:00 and 12:00 a.m.) after an overnight fast. They underwent medical assessments including an interview, blood draw and anthropometric measurements before having a one hour break for breakfast. MRI scanning was done after further psychological tests (for details see @Prehn_2020)).

@@

MRI images were retrieved with a 12-channel head coil of a 3 Tesla Trio, Siemens (Erlangen) with syngo B17 software.
<!-- MRI --> T1-weighted anatomical images were acquired as described in @Prehn_2020 (with MPRAGE, repetition time (TR) = 1900 ms, echo time (TE) = 2.52 ms, flip angle = 9°, voxel size = 1 x 1 x 1 mm$^3$, 192 sagital slices).
<!-- rsfMRI --> Resting-state echo-planar imaging  was acquired with a TR of 2.3 s and TE of 30 ms. The image matrix was 64 x 64 with an in-plane resolution of 3 mm x 3 mm and 34 slices with a slice thickness of 4 mm. 150 volumes were acquired, resulting in a total acquisition time of 5:45 minutes. Additionally, a gradient echo field map with a TE difference of 2.46 ms was acquired to correct for field inhomogeneities.
Participants were instructed to close their eyes but to remain awake during scanning.

@@

### Anatomical preprocessing
Within the module for anatomical preprocessing, the T1-weighted images were first processed by FreeSurfer's cross-sectional pipeline [@fs_cross]. Then, Freesurfer's longitudinal stream was applied to all cross-sectional runs [@fs_long]. Here, white matter (WM) and cerebral spinal fluid (CSF) masks were derived based on FreeSurfer's `aseg.mgz` segmentation file for quality control of rs preprocessing.  
The skull-stripped brain (`brainmask.mgz`) was then coregistered to the MNI152 1x1x1x mm template using ANTS [@avants2009ants]. The following parameters were used for rigid [metric: mutual information, number of iterations: [1000, 500, 250, 100]], affine [metric: mutual information, number of iterations: [1000, 500, 250, 100]] and SyN: [metric: cross-correlation, iterations: [100, 70, 50, 20]] registration steps with shrink factors [8, 4, 2, 1], and smoothing sigmas=[3, 2, 1, 0] for each step.

@@

### Minimal functional preprocessing
The function preprocessing module included the removal of first four volumes, motion correction (FSL's `MCFLIRT`), fieldmap distortion correction (FSL's `fsl_prepare_fieldmap` and `FUGUE`) and coregistration to the subject’s individual longitudinal anatomical space (FreeSurfer's `bbregister`). In more detail, the transformations derived from the latter three steps were combined into one and applied in a single step. 
For further analysis and ICA-AROMA processing, the minimally preprocessed data were intensity normalized and smoothed with a 6mm Gaussian kernel (`fslmaths -kernel gauss 2.548`).

## Denoising pipelines
Among several denoising pipelines, ICA-AROMA provides an acceptable trade-off between the mitigation of motion-FC correlation and the introduction of a distance-dependence on motion-FC relation-ship [@Parkes_2018]. @Ciric_2017 showed that GSR is very efficient in re-moving the correlation of head motion and FC, and outperforms ICA-AROMA alone. Yet, it introduces spatial dependency and spurious correlations. Thus, we pre-registered the main analyses to be performed with these two denoising pipelines.

@@

### Functional connectivity (FC)
To derive RN and DMN FC maps, we used nucleus accumbens (NAcc) and precu-neus as seed regions of interest (ROI), respectively. We did not select vmPFC for the RN due to low SNR. Because we knew about the large amount of motion in this dataset beforehand, we decided against ICA to construct the networks, as ICA is recommended to work on minimally preprocessed data. Based on FreeSurfer’s aseg.mgz and Desikan-Killiany parcellation, we created seed masks using mri_binarize (thresholded for NAcc at 26,58; precuneus at 1025,2025) and averaged them over hemispheres. Then, we used NiftiLa-belsMasker and NiftiMasker to extract the standardized timeseries from the seed regions and the whole brain. We calculated the Pearson’s correlation between them with numpy.dot, performed r-to-z Fisher-transformation and saved the resulting correlation maps for each processing mode (minimally preprocessed, aro-ma+cc, aroma+gsr). Finally, the connectivity maps were transformed into MNI space using the affine transformation and non-linear warp derived with ANTS during anatomical preprocessing.

@@

#### RsfMRI quality control
RsfMRI quality control was performed according to the protocol by @Ciric_2018. 
Head motion is a major confounder in rsfMRI. We calculated mean framewise displacement (mFD) according to @Power_2012, and log-transformed it for further analysis (logmFD).
As pre-registered, we did not exclude anybody based on high average head motion [@Beyer_2020]. Instead, we applied best-practice motion correction (AROMA+CC and AROMA+CC+GSR) and monitored the reduction of motion-related artifact according to current recommendations [@Ciric_2018].

@@

##### Group level resting-state rsfMRI QC

As described in @Beyer_2020, there was a group-by-time interaction on mFD (see Figure \@ref(fig:figDvarsmFD), upper right panel).  The intervention group reduced in head motion in the follow-up assessments compared to the control group. 
For this study, we further explored mFD and its relation to DVARS, the root mean square of the temporal change of the voxel-wise signal at each time point.
```{r figDvarsmFD, out.width="100%", fig.cap='Summary of quality measures DVARS and mFD', fig.show='hold',fig.align='center', echo=FALSE}
figList <- readRDS(file = "../report/fig/figDvarsmFDList.rds")
#(figList$fig1 | figList$fig2) / figList$fig3

@@

### Aggregated FC values
To extract aggregated FC values, we first calculated the mean DMN and RN over all participants and NAccs, adjusted for age and sex. We used GSR regressed data as input and clusterwise bootstrapping with $N = 1000$. Network masks were formed from all voxels within clusters which survived a clusterwise multiple comparison correction of FWE-corrected $p < 0.05$. We extracted the average FC from these masks for GSR-data.

@@

<!-- MODEL SETUP --> 
#### Confirmatory Analysis
We tested the pre-registered hypothesis of a  time by group interaction:  $FC=group + time + group * time$. 
As pre-registered, we performed the analysis for baseline and followup time points only, in both denoising pipelines and adjusting for no confounders (CA 1), age, sex and average logmFD of both timepoints (CA2) and age, sex, average logmFD and baseline BMI (CA3). 
In this model, time was represented as factor because we took into account possible non-linear time courses in the increase and decrease of FC over the course of one year, which may occur depending on the phase of weight management [@Olivo_2017]. The resulting factorial design contained one regressor for each time point per group (see supplements for depiction of design matrix).

#### Exploratory analysis
##### Group-by-time interaction in three time points
Then, we extended our whole-brain analysis of the group-by-time point interaction to the data from all three timepoints. This analysis had not been pre-registered. Here, we used individual logmFD values (not averaged) as covariate to capture variance in logmFD change over time points.We investigated two models (EA1.1: group x time, age, sex and EA1.2: group x time, age, sex and logmFD).

##### Average and change BMI
We further  explored the relationship of average and change in BMI in a separate group of models. As proposed by @Guillaume_2014, we calculated the between- and within subject centered values of BMI. This model, containing average BMI and BMI variability, allowed us to disentangle the differential effects of overall BMI and change in BMI on FC. We first estimated their effects in a model adjusting for baseline age and sex (Model EA2.1) and then additionally controlling for log mFD (Model EA 2.2). As we previously reported correlated change in BMI and mFD in this sample [@Beyer_2019], we explored a refined model including average BMI and logmFD and change in both measures, along with baseline age and sex (Model EA 2.3c). Here, we aimed to see whether any effect of change in BMI would be detectable when adjusting for the change in head motion. 
For a better understanding of the unique contribution of average and longitudinal change FD measures, we separately employed a third post-hoc, exploratory model (EA 2.4), solely including the head motion measures as covariates of interest: $FC = between-subject logmFD + within-subject logmFD$ age and sex as nuisance covariates.

@@

### Model input and brain mask
For each brain network, we examined the models for the two denoising pipelines (AROMA+CC/ AROMA+CC+GSR)as pre-registered. For all analyses, we used an explicit brain mask, which was a whole brain derived from the MNI ICBM “152 nonlinear 6th generation” atlas (re-sampled to 3 x 3 x 3 mm$^3$ and thresholded at 0.5 GM probability).

@@

### Model CA1 (without adjustments)
Against our initial hypothesis, there was no interaction effect of group and time point for FC for neither the NAcc nor the PCC seeds. There also was no significant main effect for any of the covariates of interest (time, group) in clusterwise inference with FWE-correction.

@@

###  Model CA3 (adjusted for age, sex, average of logmFD and baseline BMI)
There was no significant interaction when adding baseline BMI.

@@

##### Average and change in BMI, adjusted for logmFD (EA 2.1 and EA 2.2)
NAcc connectivity was not significantly related to neither mean nor change in BMI, for either of the preprocessing schemes and regardless whether we adjusted for logmFD.

>>>>

---
title: "Effects of bariatric surgery on resting-state functional connectivity of the reward and default mode network: a pre-registered analysis"
author:
- Hannah Sophie Heinrichs* [1]
- Frauke Beyer* [1][2]
- Evelyn Medawar [1]
- Kristin Prehn [3][4]
- Jürgen Ordemann [7][8]
- Agnes Floel [5][6]
- A. Veronica Witte [1][2][9]
date: 1 Max-Planck-Institute for Human Cognitive and Brain Sciences, Leipzig \newline
  2 CRC 1052 "Obesity Mechanisms", Subproject A1, University of Leipzig \newline 3 Department of Neurology and NeuroCure Clinical Research Center, Charité University Medicine, Berlin \newline 4 Department of Psychology, Medical School Hamburg, Hamburg \newline 5 Department of Neurology, University of Greifswald, Greifswald \newline 6 German Center for Neurodegenerative Diseases, Standort Rostock/Greifswald, Greifswald \newline 7 Center for Bariatric and Metabolic Surgery, Charité University Medicine, Berlin \newline 8 Center for Bariatric and Metabolic Surgery, Vivantes Clinic Spandau, Berlin \newline 9 Day Clinic for Cognitive Neurology, University Clinic Leipzig
output:
  bookdown::pdf_book:
    extra_dependencies: "subfig"
    fig_caption: yes
    number_sections: yes
    toc: yes
bibliography: "bibliography.bib"
editor_options:
  chunk_output_type: console
header-includes:
- \usepackage{setspace}
- \onehalfspacing
indent: true
---

@@

Obesity imposes serious health risks and relates to differences in resting-state functional connectivity (FC) of brain networks involved in eating behavior. Bariatric surgery is an effective treatment, but its effects on FC are still under debate. In this pre-registered analysis, we aimed to determine the effects of bariatric surgery on FC in the reward (RN) and default mode network (DMN). 
33 bariatric surgery patients and 15 obese waiting-list control patients (37 females; aged 44.15 ± 11.86 SD years (range 21-68)) underwent magnetic resonance imaging at baseline, after 6 and 12 months. We pre-registered (https://osf.io/f8tpn/https://osf.io/59bh7/) a whole-brain time-by-group interaction analysis, and a time-by-group interaction analysis on within-network FC. In exploratory analyses, we investigated all three time points and assessed the effects of weight loss and head motion. 
Bariatric surgery compared to waiting did not significantly affect FC of the RN and DMN (FWE p > .05). There also was no interaction effect on within-network FC. In exploratory analyses, surgery-related BMI decrease (FWE p= .041) and higher average head motion (FWE p= .021) resulted in significantly stronger FC in medial posterior frontal regions within the RN. 
This pre-registered analysis of longitudinal neuroimaging data did not support a strong effect of bariatric surgery, compared to waiting, on RN and DMN FC after 6 months. Exploratory analyses indicated that differences in head motion might have confounded the effects, thus a more rigorous control of within-scanner head motion during data acquisition is needed to better understand the effects of bariatric surgery on brain organization.

@@

Obesity is a worldwide health issue, entailing huge personal and societal costs. Excess amount of body fat not only affects cardiovascular and metabolic health, but also increases the risk for cognitive decline and dementia later in life [@Albanese_2017]. Conservative treatment options including behavioral therapy often do not yield the desired weight loss, especially in patients with very high BMI ($> 35$ kg/m$^2$). Here, bariatric surgery is a viable option to rapidly induce weight loss and improve glycemic status. RYBG is the most invasive, yet most effective surgical pro-cedure, but other techniques like vertical sleeve gastrectomy (VSG) and gastric banding (GB) are also common [@Berthoud_2011]. 
Bariatric surgery leads to weight loss due to malabsorption of nutrients, reduced digestion efficiency and altered food perception, appetite and central regulation of food intake [@Mulla_2017; @Brutman_2019]. Precise mechanisms how bariatric surgery leads to altered appetitive signaling are yet to be elucidated. 
Resting-state functional magnetic resonance imaging (rsfMRI) is a technique to investigate the dynamic organization of the brain. Functional connectivity (FC) networks, i.e. brain regions with correlated neural activity over time, are in anatomical correspondence with specific brain networks involved in cognitive processes, including attention and executive control [@Smith_2009]. The reward network (RN), processing hedonic value and internal motivation, and the default mode network (DMN), a higher-order network, involved in interoception and governing shifts between external-internal processes,  are promising candidates to mediate altered central regulation of food intake after bariatric surgery.
The RN (RN), comprises the ventromedial prefrontal cortex (vmPFC), the nucleus accumbens (Nacc), the putamen, the amygdala and the anterior insula  [@Odoherty_2004;@Liu_2011]. These brain regions have been found to guide food valuation processes and decision-making in humans [@Bartra_2013; @Hare_2011; @Hutcherson_2012; @Schmidt_2018]. Frequently, obesity has been associated with hyperactivation of RN regions during anticipation of (high-caloric) food cues, and in contrast, reduced activation to actual taste of these foods [@Devoto_2018; @Garcia_2014; @Meng_2020; @Stoeckel_2009; though see @Morys_2020]. RsfMRI studies also showed increased local FC of reward network regions, i.e. Nacc, vmPFC, putamen, insula [@Contreras_2017; @Coveleskie_2015;@Hogenkamp_2016], and altered connectivity with salience, homeostatic and sensorimotor networks [@Lips_2014;  @Wijngaarden_2015]. In bariatric surgery patients, connectivity  within the RN (e.g. putamen, OFC) might be normalized by the surgery, however the evidence is limited due to a lack of longitudinal obese control groups  [@Duan_2020; @Wiemerslage_2016]. Possibly, a reconfiguration of the fronto-striatal brain networks could emerge from altered gut signaling , e.g. changes in ghrelin levels, via hypothalamic-striatal projections [@Li_2019; @Karra_2013; but @Zoon_2018] . 
The DMN includes the posterior cingulate cortex (PCC)/precuneus, the medial prefrontal cortex (mPFC) and the inferior and lateral parietal cortex [@Raichle_2015]. Higher BMI and obesity have been associated with a pattern of decreased within DMN FC and increased FC of DMN regions to other networks, i.e. salience and sensory networks in several resting-state and task-based rsfMRI studies [@Beyer_2017; @Kullmann_2011; @Chao_2018; @Wijngaarden_2015; @Borowitz_2020; @Ding_2020; @Doucet_2017; @Sadler_2018]. After bariatric surgery, a normalization of the connectivity between DMN and cognitive control and salience brain regions might occur, yet no study has included a longitudinal control group [@Li_2018; @Olivo_2017; @Frank_2013].

@@

Further, while higher BMI has been consistently associated with more head motion during rsfMRI, previous studies in bariatric surgery patients have not taken this important confounder of FC into account [@Beyer_2017;@Hodgson_2016]. In the present sample, we previously reported a group-by-time interaction on head motion [@Beyer_2020]. Thus, we aimed to rigorously control for motion-related variance in our analyses. 
We had the following confirmatory hypotheses:

@@

Participants were recruited from the Center for Bariatric and Metabolic Surgery at the Charité University Medicine Berlin. Inclusion criteria were, in accordance with German guidelines for bariatric surgery, a failure of conservative obesity treatment and either (1) a BMI $> 40$ km/m$^2$ or (2) a BMI $> 35$ kg/m$^2$ and at least one typical co-morbidity (e.g. type-2 diabetes, hypertension, non-alcoholic fatty liver disease)[@Mechanick_2013]. Participants were aged between $18$ and $70$ years and had no history of cancer, chronic inflammatory disease and addiction, other severe untreated diseases, brain pathologies identified in the MRI scan or cognitive impairments (defined as MMSE score < $34$).
 In total, $51$ participants out of the originally enrolled $69$ subjects received MRI. Five data points of three subjects had to be excluded due to bad anatomical image quality (see below), which led to a final data set with $101$ rsfMRI sessions.  
<!-- DESCRIPTIVE STATISTICS -->
The final sample entailed `r nrow(subj_df)` morbidly obese individuals (`r nrow(filter(subj_df,Sex == "f"))` females; aged `r round(mean(subj_df$Age_BL),2)` $\pm$ `r round(sd(subj_df$Age_BL),2)` SD years (range `r min(subj_df$Age_BL)`-`r max(subj_df$Age_BL)`)). $X$% of participants had clinically diagnosed hypertension, $x$% had type-2 or type-1 diabetes, and $X$% smoked.
Participants either underwent surgery (n = `r cs["Sum","IG"]`, `r cs["f","IG"]` female) or were waiting list controls (n = `r cs["Sum","KG"]`, `r cs["f","KG"]` female), who waited for their health insurance's approval to undergo surgery. Measures were taken at baseline (BL), $6$ (FU1) and $12$ (FU2) months post-surgery/baseline appointment to capture both phases of rapid weight-loss and maintenance [@Maciejewski_2016]. 
Analyses were performed on all participants who provided at least one data point of rsfMRI data. `r sum(freqtable$Freq == 3)` participants had complete data, `r sum(freqtable$Freq == 2)` provided data for two time points and `r sum(freqtable$Freq == 1)` for one time point (for more details see Table S1). The pre-planned analysis of changes from baseline to follow-up included $24$ participants with both time points, in total $72$ data points.

@@

Participants arrived in the morning (between 07:00 and 12:00 a.m.) after an overnight fast. They underwent medical assessments including an interview, blood draw and anthropometric measurements before having a one hour break for breakfast. MRI scanning was done after performing a psychological test battery (for details see @Prehn_2020)).

@@

MRI was performed with a 12-channel head coil on a 3 Tesla Trio, Siemens (Erlangen) with syngo B17 software.
<!-- MRI --> T1-weighted anatomical images were acquired as described in @Prehn_2020 (with MPRAGE, repetition time (TR) = 1900 ms, echo time (TE) = 2.52 ms, flip angle = 9°, voxel size = 1 x 1 x 1 mm$^3$, 192 sagital slices).
<!-- rsfMRI --> Resting-state echo-planar imaging  was acquired with a TR of 2.3 s and TE of 30 ms. The image matrix was 64 x 64 with an in-plane resolution of 3 mm x 3 mm and 34 slices with a slice thickness of 4 mm. 150 volumes were acquired, resulting in a total acquisition time of 5:45 minutes. Additionally, a gradient echo field map with a TE difference of 2.46 ms was acquired to correct for field inhomogeneities.
Participants were instructed to close their eyes but to remain awake during scanning.

@@

### Anatomical preprocessing
Within the module for anatomical preprocessing, the T1-weighted images were first processed by FreeSurfer's cross-sectional pipeline [@fs_cross]. Then, Freesurfer's longitudinal stream was applied to all cross-sectional runs [@fs_long]. Here, white matter (WM) and cerebral spinal fluid (CSF) masks were derived based on FreeSurfer's `aseg.mgz` segmentation file for quality control of rsfMRI preprocessing.  
The skull-stripped brain (`brainmask.mgz`) was then coregistered to the MNI152 1x1x1x mm template using ANTS [@avants2009ants]. The following parameters were used for rigid [metric: mutual information, number of iterations: [1000, 500, 250, 100]], affine [metric: mutual information, number of iterations: [1000, 500, 250, 100]] and SyN: [metric: cross-correlation, iterations: [100, 70, 50, 20]] registration steps with shrink factors [8, 4, 2, 1], and smoothing sigmas=[3, 2, 1, 0] for each step.

@@

### Minimal functional preprocessing
The functional preprocessing included the removal of first four volumes, motion correction (FSL's `MCFLIRT`), fieldmap distortion correction (FSL's `fsl_prepare_fieldmap` and `FUGUE`) and coregistration to the subject’s individual longitudinal anatomical space (FreeSurfer's `bbregister`). In more detail, the transformations derived from the latter three steps were combined into one and applied in a single step. 
For further analysis and ICA-AROMA processing, the minimally preprocessed data were intensity normalized and smoothed with a 6mm Gaussian kernel (`fslmaths -kernel gauss 2.548`).

## Denoising pipelines
Among several denoising pipelines, ICA-AROMA provides an acceptable trade-off between the mitigation of motion-FC correlation and the introduction of a distance-dependence on motion-FC relation-ship [@Parkes_2018]. @Ciric_2017 showed that GSR is very efficient in re-moving the correlation of head motion and FC, and outperforms ICA-AROMA alone. Yet, it introduces spatial dependency and spurious correlations. Thus, we pre-registered the main analyses to be performed with both denoising pipelines.

@@

### Functional connectivity (FC)
To derive RN and DMN FC maps, we used Nacc and precuneus as seed regions of interest (ROI), respectively. We did not select vmPFC for the RN due to low SNR. Because we knew about the large amount of motion in this dataset beforehand, we decided against ICA to construct the networks, as ICA is recommended to work on minimally preprocessed data. Based on FreeSurfer’s aseg.mgz and Desikan-Killiany parcellation, we created seed masks using mri_binarize (thresholded for Nacc at 26,58; precuneus at 1025,2025) and averaged them over hemispheres. Then, we used NiftiLabelsMasker and NiftiMasker to extract the standardized timeseries from the seed regions and the whole brain. We calculated the Pearson’s correlation between them with numpy.dot, performed r-to-z Fisher-transformation and saved the resulting correlation maps for each preprocessing pipeline (minimally preprocessed,AROMA, AROMA+CC, AROMA+GSR). Finally, the connectivity maps were transformed into MNI space using the affine transformation and non-linear warp derived with ANTS during anatomical preprocessing.

@@

#### RsfMRI quality control (QC)
RsfMRI quality control was performed according to the protocol by @Ciric_2018. 
We calculated mean framewise displacement (mFD) according to @Power_2012, and log-transformed it for further analysis (logmFD).
As pre-registered, we did not exclude anybody based on high average head motion [@Beyer_2020]. Instead, we applied best-practice motion correction (AROMA+CC and AROMA+CC+GSR) and monitored the reduction of motion-related artifact according to current recommendations [@Ciric_2018].

@@

##### Group level rsfMRI QC

As described in @Beyer_2020, there was a group-by-time interaction on mFD (see Figure \@ref(fig:figDvarsmFD), upper right panel). The intervention group reduced in head motion in the follow-up assessments compared to the control group. For this study, we further explored mFD and its relation to DVARS, the root mean square of the temporal change of the voxel-wise signal at each time point. DVARS measures volume-to-volume signal variation.
```{r figDvarsmFD, out.width="100%", fig.cap='Summary of quality measures DVARS and mFD', fig.show='hold',fig.align='center', echo=FALSE}
figList <- readRDS(file = "../report/fig/figDvarsmFDList.rds")
#(figList$fig1 | figList$fig2) / figList$fig3

@@

### Aggregated FC values
To extract aggregated FC values, we first calculated the mean DMN and RN over all participants and Naccs, adjusted for age and sex. We used GSR regressed data as input and clusterwise bootstrapping with $N = 1000$. Network masks were formed from all voxels within clusters which survived a clusterwise multiple comparison correction of FWE-corrected $p < 0.05$. We extracted the average FC from these masks for GSR-data.

@@

<!-- MODEL SETUP --> 
#### Confirmatory Analysis
We tested the pre-registered hypothesis of a time-by-group interaction:  $FC=group + time + group * time$. 
As pre-registered, we performed the analysis for baseline and followup time points only, for AROMA+CC and AROMA+CC+GSR denoising pipelines and adjusting for no confounders (CA 1), age, sex and average logmFD of both timepoints (CA2) and age, sex, average logmFD and baseline BMI (CA3). 
In this model, time was represented as factor taking into account possible non-linear time courses in the increase and decrease of FC over the course of one year, which may occur depending on the phase of weight management [@Olivo_2017]. The resulting factorial design contained one regressor for each time point per group (see supplements for depiction of design matrix).

#### Exploratory analysis
##### Group-by-time interaction in three time points
Then, we extended our whole-brain analysis of the group-by-time point interaction to the data from all three time points. This analysis had not been pre-registered. Here, we used individual logmFD values (not averaged) as covariate to capture variance in logmFD change over time points. We investigated two models (EA1.1: group x time, age, sex and EA1.2: group x time, age, sex and logmFD).

##### Average and change BMI
We further  explored the relationship of average and change in BMI in a separate group of models. As proposed by @Guillaume_2014, we calculated the between- and within subject centered values of BMI. This model, containing average BMI and BMI change, allowed us to disentangle the differential effects of these variables on FC. We first estimated their effects in a model adjusting for baseline age and sex (Model EA2.1) and then additionally controlling for logmFD (Model EA 2.2). As we previously reported correlated change in BMI and mFD in this sample [@Beyer_2019], we explored a refined model including average BMI and logmFD and change in both measures, along with baseline age and sex (Model EA 2.3c). Here, we aimed to see whether any effect of change in BMI would be detectable when adjusting for the change in head motion. 
For a better understanding of the unique contribution of average and longitudinal change in logmFD measures, we separately employed a third post-hoc, exploratory model (EA 2.4), solely including the head motion measures as covariates of interest: $FC = between-subject logmFD + within-subject logmFD$ with age and sex as nuisance covariates.

@@

### Model input and brain mask
For each brain network, we examined the models for the two denoising pipelines (AROMA+CC/AROMA+CC+GSR) as pre-registered. For all analyses, we used an explicit brain mask, which was a whole brain derived from the MNI ICBM “152 nonlinear 6th generation” atlas (re-sampled to 3 x 3 x 3 mm$^3$ and thresholded at 0.5 GM probability).

@@

### Model CA1 (without adjustments)
Against our initial hypothesis, there was no interaction effect of group and time point for FC for neither the Nacc nor the PCC seeds. There also was no significant main effect for any of the covariates of interest (time, group) in clusterwise inference with FWE-correction.

@@

###  Model CA3 (adjusted for age, sex, average of logmFD and baseline BMI)
There was no significant interaction when adding baseline BMI. Yet, we found a significant main effect of time in this model. For AROMA+CC+GSR preprocessed data, we found higher FC of the Nacc with the lateral occipital cortex at baseline compared to followup (FWE $p=0.030$), and higher FC of the PCC within the DMN to the medial ACC (FWE $p=0.046$)(see supplements).

@@

##### Average and change in BMI, adjusted for logmFD (EA 2.1 and EA 2.2)
Nacc connectivity was not significantly related to neither mean nor change in BMI, for either of the preprocessing schemes and regardless whether we adjusted for logmFD.
